---
title: "Appendix S2. Model definitions, model fitting, and model evaluation."
subtitle: Supporting information for Scheuerell et al.
output:
  pdf_document:
    highlight: haddock
    toc: yes
    number_sections: true
    toc_depth: '3'
fontsize: 11pt
geometry: margin=1in
---

```{r set_options, echo = FALSE, message = FALSE}
options(width = 100)
knitr::opts_chunk$set(message = FALSE, size="small")
set.seed(123)
def_chunk_hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def_chunk_hook(x, options)
  ifelse(options$size != "normalsize",
         ## TRUE
         paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"),
         ## FALSE
         x)
})
```

\vspace{0.2in}

This is version `r paste0('0.',format(Sys.time(), '%y.%m.%d'))`.

# Background

This appendix describes how we fit the models and evaluated their relative performances. It demonstrates how to load the fish data and environmenal covariates, specify the different models in the __JAGS__ software, and fit each one. 

All analyses require the [R software](https://cran.r-project.org/) (v3.4.3 or later) for data retrieval, data processing, and summarizing model results, and the [JAGS software](http://mcmc-jags.sourceforge.net/) (v4.2.0) for Markov chain Monte Carlo (MCMC) simulation. Please note that some of the __R__ code below may not work with older versions of __JAGS__ due to some changes in the ways that arrays are handled.

We also need a few packages that are not included with the base installation of __R__, so we begin by installing them (if necessary) and then loading them.

```{r load_pkgs, message = FALSE, warning = FALSE}
if(!require("here")) {
  install.packages("here")
  library("here")
}
if(!require("readr")) {
  install.packages("readr")
  library("readr")
}
if(!require("rjags")) {
  install.packages("rjags")
  library("rjags")
}
if(!require("loo")) {
  install.packages("loo")
  library("loo")
}
## set directory locations
datadir <- here("data")
jagsdir <- here("jags")
```

We also need a couple of helper functions.

```{r define_funcs}
## better round
Re2prec <- function(x, map = "round", prec = 1) {
  ## 'fun' can be "round", "floor", or "ceiling"
  ## 'prec' is nearest value
  ## (eg, 0.1 is to nearest tenth; 1 is to nearest integer)
  if(prec<=0) { stop("\"prec\" cannot be less than or equal to 0") }
  do.call(map,list(x/prec))*prec
}

## wrapper function to fit a JAGS model
fit_jags <- function(model, data, params, inits, ctrl, dir = jagsdir) {
  jm <- jags.model(file.path(jagsdir, model),
                   data,
                   inits,
                   n.chains = ctrl$chains,
                   n.adapt = 0,
                   quiet = TRUE)
  adp <- FALSE
  while(!adp) {
    adp <- adapt(jm, n.iter = 1000)
  }
  update(jm, ctrl$burn, progress.bar = "none")
  return(coda.samples(jm, params, ctrl$length, ctrl$thin))
}

## inits function for base model
init_vals_AR <- function() {
	list(alpha = 5,
	     beta_inv = exp(mean(ln_dat_esc, na.rm = TRUE)),
	     pi_tau = 10,
	     pi_eta = rep(1,A),
	     pi_vec = matrix(c(0.05,0.5,0.4,0.05),
	                     n_yrs-age_min, A, 
	                     byrow = TRUE),
	     Rec_mu = log(1000),
	     Rec_sig = 0.1,
	     sigma_r = 0.5,
	     sigma_s = 0.1,
	     tot_ln_Rec = rep(log(1000), n_yrs - age_min),
	     innov_1 = 0,
	     phi = 0.5)
}

## inits function for cov models
init_vals_cov <- function() {
  list(alpha = 5,
       beta_inv = exp(mean(ln_dat_esc, na.rm = TRUE)),
       gamma = 0,
       pi_tau = 10,
       pi_eta = rep(1,A),
	     pi_vec = matrix(c(0.05,0.5,0.4,0.05),
                       n_yrs-age_min, A, 
                       byrow = TRUE),
       Rec_mu = log(1000),
       Rec_sig = 0.1,
       tot_ln_Rec = rep(log(1000), n_yrs - age_min),
       # phi = 0.5,
       innov_1 = 0)
}

## estimate LOOIC
looic <- function(jags_obj, mcmc_ctrl) {
  ## convert mcmc.list to matrix
  tmp_lp <- as.matrix(jags_obj)
  ## extract pointwise likelihoods
  tmp_lp <- tmp_lp[,grepl("lp_", colnames(tmp_lp))]
  ## if numerical underflows, convert -Inf to 5% less than min(likelihood)
  if(any(is.infinite(tmp_lp))) {
    tmp_lp[is.infinite(tmp_lp)] <- NA
    tmp_min <- min(tmp_lp, na.rm = TRUE)
    tmp_lp[is.na(tmp_lp)] <- tmp_min * 1.05
  }
  ## effective sample size
  r_eff <- relative_eff(exp(tmp_lp),
                        chain_id = rep(seq(mcmc_ctrl$chains),
                                       each = mcmc_ctrl$length / mcmc_ctrl$thin))
  ## calculate LOOIC
  looic <- loo(tmp_lp, r_eff = r_eff)
  return(looic)
}
```

# User inputs

We begin by supplying values for the minimum and maximum ages of spawning adults, plus some information for the model code and evaluation.

```{r get_user_inputs}
## min & max adult age classes
age_min <- 3
age_max <- 6

## file where to save JAGS model
fn_jags <- "Willamette_Chin_SR_flow_models_mainstem_JAGS.txt"

## upper threshold for Gelman & Rubin's potential scale reduction factor (Rhat).
Rhat_thresh <- 1.1
```

Next we specify the names of five necessary data files containing the following information:
 
 1. observed total number of adult spawners (escapement) by year;
 2. observed age composition of adult spawners by year;
 3. observed total harvest by year;
 4. flow covariates by year;
 5. metadata for flow covariates.


```{r get_filenames}
## 1. file with escapement data
## [n_yrs x 2] matrix of obs counts; 1st col is calendar yr
fn_esc <- "chin_esc.csv"

## 2. file with age comp data
## [n_yrs x (1+A)]; 1st col is calendar yr
fn_age <- "chin_agecomp.csv"

## 3. file with harvest data
## [n_yrs x 2] matrix of obs catch; 1st col is calendar yr
fn_harv <- "chin_harv.csv"

## 4. file with harvest data
## [n_yrs x 2] matrix of obs catch; 1st col is calendar yr
fn_cov <- "Willamette_Chin_SR_mainstem_flow_covariates.csv"

## 5. covariate metadata
cov_meta_file <- "chin_cov_metadata.csv"
```

# Loading the fish data

Here we load in the first three data files and do some simple calculations and manipulations.

First the spawner data:

```{r get_escapement_data}
## escapement
dat_esc <- read.csv(file.path(datadir, fn_esc))
## use total counts
dat_esc <- dat_esc[dat_esc$group=="total",-1]
## years of data
dat_yrs <- dat_esc$year
## number of years of data
n_yrs <- length(dat_yrs)
## get first & last years
yr_frst <- min(dat_yrs)
yr_last <- max(dat_yrs)
## log of escapement
ln_dat_esc <- log(dat_esc[,-1])
```

Next the age composition data:

```{r get_age_data}
## age comp data
dat_age <- read.csv(file.path(datadir, fn_age))
## drop first age_min rows; drop site & year col
dat_age <- dat_age[-(1:(age_min)), -1]
## num of age classes
A <- age_max-age_min+1
## total num of age obs by cal yr
dat_age[,"sum"] <- apply(dat_age,1,sum)
## row indices for any years with no obs age comp
idx_NA_yrs <- which(dat_age$sum<A, TRUE)
if(length(idx_NA_yrs) > 0) {
  ## replace 0's in yrs w/o any obs with NA's
  dat_age[idx_NA_yrs,(1:A)] <- NA
  ## change total in yrs w/o any obs from 0 to A to help dmulti()
  dat_age[idx_NA_yrs,"sum"] <- A
}
## convert class
dat_age <- as.matrix(dat_age)
```

And then the harvest data:

```{r get_harvest}
## harvest
dat_harv <- read.csv(file.path(datadir, fn_harv))
## trim to correct years & drop year col 
dat_harv <- dat_harv[dat_harv$year>=yr_frst & dat_harv$year<=yr_last,-1]
```

# Loading the covariates

Load the metadata file containing all of the specifications for the covariates to be used.

```{r get_cov_meta}
cov_meta <- read.csv(file.path(datadir, cov_meta_file), stringsAsFactors = FALSE)
cov_meta$code <- gsub("\"","",cov_meta$code)
cov_meta$begin <- gsub("\"","",cov_meta$begin)
cov_meta$end <- gsub("\"","",cov_meta$end)
```

Load the saved covariates.

```{r get_flow_cov}
cov_flow <- read.csv(file.path(datadir, fn_cov))[,-1]
n_cov <- dim(cov_flow)[2]
```


# Specifying the models in JAGS

Now we can specify the various models in JAGS. We fit a total of 4 different models, which we outline below, based on the 2 different process models with and without and covariates.

## Ricker model without covariates

```{r JAGS_RK_AR}
cat("

model {
  
  ##--------
  ## PRIORS
  ##--------
  ## alpha = exp(a) = intrinsic productivity
  alpha ~ dnorm(0,0.01) T(0,);
  mu_Rkr_a <- log(alpha);
  E_Rkr_a <- mu_Rkr_a + sigma_r/(2 - 2*phi^2);
  
  ## strength of dens depend
  beta_inv ~ dnorm(0, 1e-9) T(0,);
  beta <- 1/beta_inv;
  
  ## AR(1) coef for proc errors
  phi ~ dunif(-0.999,0.999);
  
  ## process variance for recruits model
  sigma_r ~ dnorm(0, 2e-2) T(0,);
  tau_r <- 1/sigma_r;
  
  ## innovation in first year
  innov_1 ~ dnorm(0,tau_r*(1-phi*phi));
  
  ## obs variance for spawners
  tau_s <- 1/sigma_s;
  sigma_s ~ dnorm(0, 0.001) T(0,);
  
  ## maturity schedule
  ## unif vec for Dirch prior
  theta <- c(2,20,20,1)
  ## hyper-mean for maturity
  pi_eta ~ ddirch(theta);
  ## hyper-prec for maturity
  pi_tau ~ dnorm(0, 0.01) T(0,);
  for(t in 1:(n_yrs-age_min)) { pi_vec[t,1:A] ~ ddirch(pi_eta*pi_tau) }

  ## unprojectable early recruits;
  ## hyper mean across all popns
  Rec_mu ~ dnorm(0,0.001);
  ## hyper SD across all popns
  Rec_sig ~ dunif(0,100);
  ## precision across all popns
  Rec_tau <- pow(Rec_sig,-2);
  ## multipliers for unobservable total runs
	ttl_run_mu ~ dunif(1,5);
	ttl_run_tau ~ dunif(1,20);

  ## get total cal yr returns for first age_min yrs
  for(i in 1:(age_min)) {
		ln_tot_Run[i] ~ dnorm(ttl_run_mu*Rec_mu,Rec_tau/ttl_run_tau);
		tot_Run[i] <- exp(ln_tot_Run[i]);
  }
  
  ## estimated harvest rate
  for(t in 1:n_yrs) { h_rate[t] ~ dunif(0,1) }

  ##------------
  ## LIKELIHOOD
  ##------------
  ## 1st brood yr requires different innovation
  ## predicted recruits in BY t
  ln_Rkr_a[1] <- mu_Rkr_a;
  E_ln_Rec[1] <- ln_Rkr_a[1] + ln_Sp[1] - beta*Sp[1] + phi*innov_1;
  tot_ln_Rec[1] ~ dnorm(E_ln_Rec[1],tau_r);
  res_ln_Rec[1] <- tot_ln_Rec[1] - E_ln_Rec[1];
  ## median of total recruits
  tot_Rec[1] <- exp(tot_ln_Rec[1]);
  
  ## R/S
  ln_RS[1] <- tot_ln_Rec[1] - ln_Sp[1];
  
  ## brood-yr recruits by age
  for(a in 1:A) {
    Rec[1,a] <- tot_Rec[1] * pi_vec[1,a];
  }
  
  ## brood years 2:(n_yrs-age_min)
  for(t in 2:(n_yrs-age_min)) {
    ## predicted recruits in BY t
    ln_Rkr_a[t] <- mu_Rkr_a; 
    E_ln_Rec[t] <- ln_Rkr_a[t] + ln_Sp[t] - beta*Sp[t] + phi*res_ln_Rec[t-1];
    tot_ln_Rec[t] ~ dnorm(E_ln_Rec[t],tau_r);
    res_ln_Rec[t] <- tot_ln_Rec[t] - E_ln_Rec[t];
    ## median of total recruits
    tot_Rec[t] <- exp(tot_ln_Rec[t]);
    ## R/S
    ln_RS[t] <- tot_ln_Rec[t] - ln_Sp[t];
    ## brood-yr recruits by age
    for(a in 1:A) {
      Rec[t,a] <- tot_Rec[t] * pi_vec[t,a];
    }
  } ## end t loop over year
  
  ## get predicted calendar year returns by age
  ## matrix Run has dim [(n_yrs-age_min) x A]
  ## step 1: incomplete early broods
  ## first cal yr of this grp is first brood yr + age_min
  for(i in 1:(age_max-age_min)) {
    ## projected recruits
    for(a in 1:i) {
      Run[i,a] <- Rec[i-a+1,a];
    }
    ## imputed recruits
    for(a in (i+1):A) {
      lnRec[i,a] ~ dnorm(Rec_mu,Rec_tau);
      Run[i,a] <- exp(lnRec[i,a]);
    }
    ## total run size
    tot_Run[i+age_min] <- sum(Run[i,1:A]);
    ## predicted age-prop vec for multinom
    for(a in 1:A) {
      age_v[i,a] <- Run[i,a] / tot_Run[i+age_min];
    }
    ## multinomial for age comp
    dat_age[i,1:A] ~ dmulti(age_v[i,1:A],dat_age[i,A+1]);
    lp_age[i] <- logdensity.multi(dat_age[i,1:A],age_v[i,1:A],dat_age[i,A+1]);
  }
  
  ## step 2: info from complete broods
  ## first cal yr of this grp is first brood yr + age_max
  for(i in A:(n_yrs-age_min)) {
    for(a in 1:A) {
      Run[i,a] <- Rec[i-a+1,a];
    }
    ## total run size
    tot_Run[i+age_min] <- sum(Run[i,1:A]);
    ## predicted age-prop vec for multinom
    for(a in 1:A) {
      age_v[i,a] <- Run[i,a] / tot_Run[i+age_min];
    }
    ## multinomial for age comp
    dat_age[i,1:A] ~ dmulti(age_v[i,1:A],dat_age[i,A+1]);
    lp_age[i] <- logdensity.multi(dat_age[i,1:A],age_v[i,1:A],dat_age[i,A+1]);
  }
  
  ## get predicted calendar year spawners
  ## first cal yr is first brood yr
  for(t in 1:n_yrs) {
    ## obs model for spawners
    # Sp[t] <- max(10,tot_Run[t] - dat_harv[t]);
    # est_harv[t] = h_rate[t] * tot_Run[t];
    # dat_harv[t] ~ dlnorm(log(est_harv[t]), 20);
    Sp[t] = tot_Run[t] - dat_harv[t];
    ln_Sp[t] <- log(Sp[t]);
    ln_dat_esc[t] ~ dnorm(ln_Sp[t], tau_s);
    lp_esc[t] <- logdensity.norm(ln_dat_esc[t],ln_Sp[t], tau_s);
  }
  
} ## end model description

", file=file.path(jagsdir, "IPM_RK_AR.txt"))
```

## Ricker model with covariates

```{r JAGS_RK_cov_AR}
cat("

model {
  
  ##--------
  ## PRIORS
  ##--------
  ## alpha = exp(a) = intrinsic productivity
  alpha ~ dnorm(0,0.01) T(0,);
  mu_Rkr_a <- log(alpha);
  E_Rkr_a <- mu_Rkr_a + sigma_r/(2 - 2*phi^2);
  
  ## strength of dens depend
  beta_inv ~ dnorm(0, 1e-9) T(0,);
  beta <- 1/beta_inv;
  
  ## covariate effect
  gamma ~ dnorm(0,0.01)

  ## AR(1) coef for proc errors
  # phi ~ dunif(-0.999,0.999);
  phi <- 0;
  
  ## process variance for recruits model
  sigma_r ~ dnorm(0, 2e-2) T(0,);
  tau_r <- 1/sigma_r;
  
  ## innovation in first year
  innov_1 ~ dnorm(0,tau_r*(1-phi*phi));
  
  ## obs variance for spawners
  tau_s <- 1/sigma_s;
  sigma_s ~ dnorm(0, 0.001) T(0,);
  
  ## maturity schedule
  ## unif vec for Dirch prior
  theta <- c(2,20,20,1)
  ## hyper-mean for maturity
  pi_eta ~ ddirch(theta);
  ## hyper-prec for maturity
  pi_tau ~ dnorm(0, 0.01) T(0,);
  for(t in 1:(n_yrs-age_min)) { pi_vec[t,1:A] ~ ddirch(pi_eta*pi_tau) }

  ## unprojectable early recruits;
  ## hyper mean across all popns
  Rec_mu ~ dnorm(0,0.001);
  ## hyper SD across all popns
  Rec_sig ~ dunif(0,100);
  ## precision across all popns
  Rec_tau <- pow(Rec_sig,-2);
  ## multipliers for unobservable total runs
	ttl_run_mu ~ dunif(1,5);
	ttl_run_tau ~ dunif(1,20);

  ## get total cal yr returns for first age_min yrs
  for(i in 1:(age_min)) {
		ln_tot_Run[i] ~ dnorm(ttl_run_mu*Rec_mu,Rec_tau/ttl_run_tau);
		tot_Run[i] <- exp(ln_tot_Run[i]);
  }
  
  ## estimated harvest rate
  for(t in 1:n_yrs) { h_rate[t] ~ dunif(0,1) }

  ##------------
  ## LIKELIHOOD
  ##------------
  ## 1st brood yr requires different innovation
  ## predicted recruits in BY t
  covar[1] <- gamma * mod_cvrs[1];
  ln_Rkr_a[1] <- mu_Rkr_a + covar[1]; 
  E_ln_Rec[1] <- ln_Rkr_a[1] + ln_Sp[1] - beta*Sp[1] + phi*innov_1;
  tot_ln_Rec[1] ~ dnorm(E_ln_Rec[1],tau_r);
  res_ln_Rec[1] <- tot_ln_Rec[1] - E_ln_Rec[1];
  ## median of total recruits
  tot_Rec[1] <- exp(tot_ln_Rec[1]);
  
  ## R/S
  ln_RS[1] <- tot_ln_Rec[1] - ln_Sp[1];
  
  ## brood-yr recruits by age
  for(a in 1:A) {
    Rec[1,a] <- tot_Rec[1] * pi_vec[1,a];
  }
  
  ## brood years 2:(n_yrs-age_min)
  for(t in 2:(n_yrs-age_min)) {
    ## predicted recruits in BY t
    covar[t] <- gamma * mod_cvrs[t];
    ln_Rkr_a[t] <- mu_Rkr_a + covar[t]; 
    E_ln_Rec[t] <- ln_Rkr_a[t] + ln_Sp[t] - beta*Sp[t] + phi*res_ln_Rec[t-1];
    tot_ln_Rec[t] ~ dnorm(E_ln_Rec[t],tau_r);
    res_ln_Rec[t] <- tot_ln_Rec[t] - E_ln_Rec[t];
    ## median of total recruits
    tot_Rec[t] <- exp(tot_ln_Rec[t]);
    ## R/S
    ln_RS[t] <- tot_ln_Rec[t] - ln_Sp[t];
    ## brood-yr recruits by age
    for(a in 1:A) {
      Rec[t,a] <- tot_Rec[t] * pi_vec[t,a];
    }
  } ## end t loop over year
  
  ## get predicted calendar year returns by age
  ## matrix Run has dim [(n_yrs-age_min) x A]
  ## step 1: incomplete early broods
  ## first cal yr of this grp is first brood yr + age_min
  for(i in 1:(age_max-age_min)) {
    ## projected recruits
    for(a in 1:i) {
      Run[i,a] <- Rec[i-a+1,a];
    }
    ## imputed recruits
    for(a in (i+1):A) {
      lnRec[i,a] ~ dnorm(Rec_mu,Rec_tau);
      Run[i,a] <- exp(lnRec[i,a]);
    }
    ## total run size
    tot_Run[i+age_min] <- sum(Run[i,1:A]);
    ## predicted age-prop vec for multinom
    for(a in 1:A) {
      age_v[i,a] <- Run[i,a] / tot_Run[i+age_min];
    }
    ## multinomial for age comp
    dat_age[i,1:A] ~ dmulti(age_v[i,1:A],dat_age[i,A+1]);
    lp_age[i] <- logdensity.multi(dat_age[i,1:A],age_v[i,1:A],dat_age[i,A+1]);
  }
  
  ## step 2: info from complete broods
  ## first cal yr of this grp is first brood yr + age_max
  for(i in A:(n_yrs-age_min)) {
    for(a in 1:A) {
      Run[i,a] <- Rec[i-a+1,a];
    }
    ## total run size
    tot_Run[i+age_min] <- sum(Run[i,1:A]);
    ## predicted age-prop vec for multinom
    for(a in 1:A) {
      age_v[i,a] <- Run[i,a] / tot_Run[i+age_min];
    }
    ## multinomial for age comp
    dat_age[i,1:A] ~ dmulti(age_v[i,1:A],dat_age[i,A+1]);
    lp_age[i] <- logdensity.multi(dat_age[i,1:A],age_v[i,1:A],dat_age[i,A+1]);
  }
  
  ## get predicted calendar year spawners
  ## first cal yr is first brood yr
  for(t in 1:n_yrs) {
    ## obs model for spawners
    # Sp[t] <- max(10,tot_Run[t] - dat_harv[t]);
    # est_harv[t] = h_rate[t] * tot_Run[t];
    # dat_harv[t] ~ dlnorm(log(est_harv[t]), 20);
    Sp[t] = tot_Run[t] - dat_harv[t];
    ln_Sp[t] <- log(Sp[t]);
    ln_dat_esc[t] ~ dnorm(ln_Sp[t], tau_s);
    lp_esc[t] <- logdensity.norm(ln_dat_esc[t],ln_Sp[t], tau_s);
  }
  
} ## end model description

", file=file.path(jagsdir, "IPM_RK_cov_AR.txt"))
```


# Fitting the models

Before fitting the model in JAGS, we need to specify:

1. the data and indices that go into the model;
2. the model parameters and states that we want JAGS to return;
3. the MCMC control parameters.

```{r jags_setup}
## 1. Data to pass to JAGS:
dat_jags <- list(dat_age = dat_age,
                 ln_dat_esc = ln_dat_esc,
                 dat_harv = dat_harv,
                 A = A,
                 age_min = age_min,
                 age_max = age_max,
                 n_yrs = n_yrs) 

## 2. Model params/states for JAGS to return:
par_jags <- c("alpha","E_Rkr_a","ln_Rkr_a",
              "beta",
              "Sp","Rec","tot_ln_Rec","ln_RS",
              "pi_eta","pi_tau",
              "sigma_r","sigma_s","res_ln_Rec",
              "lp_age","lp_esc")

## 3. MCMC control params:
mcmc_ctrl <- list(
  chains = 4,
  length = 1.25e5,
  burn = 5e4,
  thin = 100
)
## total number of MCMC samples after burnin
mcmc_samp <- mcmc_ctrl$length*mcmc_ctrl$chains/mcmc_ctrl$thin
```

```{r start_timer, include = FALSE}
## start timer
timer_start <- proc.time() 
```

Please note that the following code takes ~60 min to run on a quad-core machine with 3.5 GHz Intel processors.

```{r fit_models, echo = TRUE, message = FALSE, warning = FALSE}
## total number of models to fit
n_mods <- 1 + n_cov

## empty list for LOOIC values
LOOIC <- vector("list", n_mods)

## fit base model (if not already saved)
if(!file.exists(file.path(jagsdir, "fit_ricker_base.rds"))) {
  mod_fit <- fit_jags("IPM_RK_AR.txt", dat_jags, par_jags, init_vals_AR, mcmc_ctrl)
  ## save results to file
  saveRDS(mod_fit, file.path(jagsdir, "fit_ricker_base.rds"))
  ## compute LOOIC
  LOOIC[[1]] <- looic(mod_fit, mcmc_ctrl)
}

## fit models with covariates
par_jags <- c(par_jags, "gamma")
for(i in seq(n_mods-1)) {
  if(!file.exists(file.path(jagsdir, paste0("fit_ricker_cov_", i, ".rds")))) {
    dat_jags$mod_cvrs <- cov_flow[,i]
    mod_fit <- fit_jags("IPM_RK_cov_AR.txt", dat_jags, par_jags,
                        init_vals_cov, mcmc_ctrl)
    ## save results to file
    saveRDS(mod_fit, file.path(jagsdir, paste0("fit_ricker_cov_", i, ".rds")))
    ## compute LOOIC
    LOOIC[[i+1]] <- looic(mod_fit, mcmc_ctrl)
  }
}
if(!file.exists(file.path(jagsdir, "LOOIC_values.rds"))) {
       saveRDS(LOOIC, file.path(jagsdir, "LOOIC_values.rds"))
} else {
       LOOIC <- readRDS(file.path(jagsdir, "LOOIC_values.rds"))
}
```

```{r stop_timer, include = FALSE}
## stop timer
run_time_in_min <- round(((proc.time()-timer_start)/60)["elapsed"], 1)
cat(run_time_in_min, file = "run_time_in_min.txt")
```

#### Convergence checks

```{r diag_Ricker}
base_mod <- readRDS(file.path(jagsdir, "fit_ricker_base.rds"))

par_conv <- c("alpha", "beta",
              "sigma_r", "sigma_s",
              "pi_tau", paste0("pi_eta[",seq(A),"]"))

## Gelman-Rubin
gelman.diag(base_mod[,par_conv])

## autocorrelation
t(round(autocorr.diag(base_mod[,par_conv],
                      lags = seq(mcmc_ctrl$thin, 4*mcmc_ctrl$thin, mcmc_ctrl$thin),
                      relative=FALSE), 2))
```

# Model selection

Via `loo()` and `compare()` with full table of results. Note that `elpd_diff` will be negative (positive) if the expected predictive accuracy for the first (second) model is higher.

```{r get_LOOIC_c, warning=FALSE}
## table of LOOIC values
tbl_LOOIC <- as.data.frame(round(compare(x = LOOIC), 1))
tbl_LOOIC$d_looic <- -2 * tbl_LOOIC$elpd_diff
tbl_LOOIC <- tbl_LOOIC[, c("p_loo", "se_p_loo", "looic", "se_looic", "d_looic")]
rownames(tbl_LOOIC) <- sub("model", "", rownames(tbl_LOOIC))
tbl_LOOIC <- tbl_LOOIC[order(as.numeric(rownames(tbl_LOOIC))),]
tbl_LOOIC <- data.frame(life_stage = c("base", cov_meta$life_stage),
                        variable = c("NA", sub(" of 7-day mean","",cov_meta$long_name)),
                        begin = c("NA", cov_meta$begin),
                        end = c("NA", cov_meta$end),
                        lag = c(NA, cov_meta$lag_1),
                        tbl_LOOIC)
tbl_LOOIC[order(tbl_LOOIC[,"looic"]), ]

## best model
best_i <- which(tbl_LOOIC[,"looic"] == min(tbl_LOOIC[,"looic"]))
best_fit <- readRDS(file.path(jagsdir, paste0("fit_ricker_cov_", best_i, ".rds")))
```


# Model diagnostics

## Gelman & Rubin statistic

Here is a table of the Gelman & Rubin statistics $(R_{hat})$ for the estimated parameters. Recall that we set an upper threshold of `r Rhat_thresh`, so values larger than that deserve some additional inspection.

```{r diag_rhat}
## params of interest
par_conv <- c("alpha","beta","gamma",
              "sigma_r","sigma_s",
              "pi_tau",paste0("pi_eta[",seq(A-1),"]"),
              paste0("Sp[",seq(n_yrs),"]"),
              paste0("tot_ln_Rec[",seq(n_yrs-age_min),"]"))
## Gelman-Rubin
gelman.diag(best_fit[,par_conv])
```

## Autocorrelation

```{r diag_acf}
t(round(autocorr.diag(best_fit[,par_conv],
                      lags = seq(mcmc_ctrl$thin, 3*mcmc_ctrl$thin, mcmc_ctrl$thin),
                      relative=FALSE), 2))
```


## Effective sample sizes

```{r diag_neff}
floor(effectiveSize(best_fit))
```


