---
title: Appendix S3. Code to recreate figures from main text.
output:
  pdf_document:
    highlight: haddock
    toc: yes
    toc_depth: '3'
fontsize: 11pt
geometry: margin=1in
---

```{r set_options, echo = FALSE, message = FALSE}
options(width = 100)
knitr::opts_chunk$set(message = FALSE, size="small", vspaceplot='2em')
def_chunk_hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def_chunk_hook(x, options)
  ifelse(options$size != "normalsize",
         ## TRUE
         paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"),
         ## FALSE
         x)
})
```

\vspace{0.2in}

This is version `r paste0('0.',format(Sys.time(), '%y.%m.%d'))`.

# Background

This appendix shows how to recreate the figures in the main text based on the results from the best of the fitted models.

All analyses require the [R software](https://cran.r-project.org/) (v3.5 or later), as well as a few packages that are not included with the base installation of R.

```{r load_pkgs, message = FALSE, warning = FALSE}
## for dir management
if(!require("here")) {
  install.packages("here")
  library("here")
}
## set directory locations
datadir <- here("data")
jagsdir <- here("jags")
figsdir <- here("report/figures")
## for reading csv
if(!require("readr")) {
  install.packages("readr")
  library("readr")
}
## for Lambert function
if(!require("gsl")) {
  install.packages("gsl")
  library("gsl")
}
## for figure captions
if(!require("captioner")) {
  devtools::install_github("adletaw/captioner")
  library("captioner")
}
## set default caption options
fig_cap <- captioner(prefix = "Figure S", suffix = ". ", auto_space = FALSE,
                     style = "b", style_prefix = TRUE)
```

We also need the following helper function.

```{r define_funcs}
Re2prec <- function(x, map = "round", prec = 1) {
  ## 'map' can be "round", "floor", or "ceiling"
  ## 'prec' is nearest value
  ## (eg, 0.1 is to nearest tenth; 1 is to nearest integer)
  if(prec <= 0) {
    stop("\"prec\" cannot be less than or equal to 0")
  }
  do.call(map, list(x / prec)) * prec
}
```

```{r get_user_inputs, echo=FALSE}
## first & last years of fish data
yr_frst <- 1999
yr_last <- 2017
## years of data
dat_yrs <- seq(yr_frst,yr_last)
## number of years of data
n_yrs <- length(dat_yrs)

## min & max adult age classes
age_min <- 3
age_max <- 6
## num of age classes
A <- age_max - age_min + 1

## number of years for run forecasts
n_fore <- 0

## posterior coverage interval
CI_vec <- c(0.025,0.5,0.975)

## covariate names & units for plotting
cov_name <- expression(paste("Flow (",ft^3," ",s^{-1},")"))

## best model
best_i <- 25
```

# Load the information

Here we load in the model fits, covariates, and harvest data.

```{r load_mod_fits}
best_fit <- readRDS(file.path(jagsdir, paste0("fit_ricker_cov_", best_i, ".rds")))
```

```{r load_covariates}
## covariate(s)
dat_cvrs <- read_csv(file.path(datadir, "Willamette_Chin_SR_mainstem_flow_covariates.csv"))
t_idx <- seq(yr_frst, length.out = n_yrs-age_min)
dat_cvrs <- as.matrix(dat_cvrs[seq(length(t_idx)),])
```

```{r get_escapement_data}
## escapement
dat_esc <- read_csv(file.path(datadir, "chin_esc.csv"))
## use total counts
dat_esc <- dat_esc[dat_esc$group=="total",-1]
## log of escapement
ln_dat_esc <- log(dat_esc[,-1])
```

```{r get_harvest}
## harvest
dat_harv <- read_csv(file.path(datadir, "chin_harv.csv"))
## trim to correct years & drop year col 
dat_harv <- dat_harv[dat_harv$year>=yr_frst & dat_harv$year<=yr_last,-1]
```

\newpage

# Extract posterior values

```{r}
## convert the `mcmc.list` output into a more user-friendly form
mod_res <- do.call("rbind", best_fit)

## spawners
s_pst <- mod_res[,grep("Sp", colnames(mod_res))]
s_pst <- apply(s_pst, 2, quantile, CI_vec)
s_dat <- s_pst[, 1:(n_yrs-age_min+n_fore)]

## total run size
run_size <- s_pst + matrix(as.matrix(dat_harv), length(CI_vec), n_yrs, byrow = TRUE)

## posterior of recruits
r_dat <- mod_res[, grep("tot_ln_Rec", colnames(mod_res))]
r_dat <- exp(apply(r_dat, 2, quantile, CI_vec))

## median values for a & b
aa <- apply(mod_res[, grep("ln_Rkr_a", colnames(mod_res))], 2, median)
bb <- median(mod_res[, "beta"])

## alpha w/ effect of median flow
alpha_est <- mod_res[, "alpha"] * exp(mod_res[, "gamma"] * median(dat_cvrs[,best_i+1]))
alphaCI <- quantile(alpha_est, CI_vec)
## pile into last ban for plotting
alpha_est[alpha_est > 5] <- 5

## carrying capacity in 1000s
K_est <- log(alpha_est) / mod_res[, "beta"] / 1000
K_est <- K_est[K_est > 0]
K_est_CI <- quantile(K_est, CI_vec)
## pile into last ban for plotting
K_est[K_est > 24] <- 24
```

\newpage

# Figures

## Deterministic Ricker model

```{r model_forms, fig.height=4, fig.width=4, fig.pos="placeHere", fig.align="center", message=FALSE}
## params
ra <- 3
rb <- 1.2e-4

## S-R curve
## spawners
ss <- seq(0,1.2e4,10)
## recruits
rr <- ra*ss/exp(rb*ss)

png(file.path(figsdir, "Fig_1_Ricker_form.png"),
     height = 4, width = 4, units = "in", res = 500)

par(mai=c(0.8,0.6,0.2,0.2), omi=c(0,0,0,0.25))
plot(ss, rr, type="n", xlim=range(ss), ylim=range(ss), xaxs="i", yaxs="i",
     xlab="", ylab="", xaxt="n", yaxt="n", bty="L")
mtext(expression(italic(S[t])), 1, line=1, cex=1.1, at=max(ss))
mtext(expression(italic(R[t])), 2, line=0.5, cex=1.1, at=max(ss), las=1)
## 1:1
abline(a=0, b=1, col="gray")
## R-S
lines(ss, rr, lwd=2)
rmod <- expression(frac(italic(alpha * S[t]),e^italic({beta * S[t]})))
text(12300, ra*max(ss)/exp(rb*max(ss)), rmod, adj=c(0,0.5), xpd=NA)
## alpha
segments(0, 0, 1900, ra*1900, lty="dashed")
text(2000, ra*2000, expression(alpha), adj=c(0.5,0.5))
## K
segments(0, log(ra)/rb, log(ra)/rb, log(ra)/rb, lty="dashed")
segments(log(ra)/rb, 0, log(ra)/rb, log(ra)/rb, lty="dashed")
text(log(ra)/rb, 0, expression(frac(log(alpha),beta)), adj=c(0.5,1.2), xpd=NA)
text(0, log(ra)/rb, expression(italic(K)), pos=2, xpd=NA)

dev.off()
```

```{r, echo=FALSE}
cnt <- 1
```

\newpage

\begin{center}
\includegraphics[width=0.8\textwidth]{figures/Fig_1_Ricker_form}
\end{center}

`r fig_cap(cnt, caption = "Deterministic form of the Ricker model used in the analysis. The parameter $\\alpha$ is the slope at the origin and the derived parameter $K$ is the carrying capacity. The gray line is where $R_t = S_t$.", display = "full")` 

\newpage

## Total population size

```{r plot_run_size, fig.width = 7, fig.height = 4, fig.align="center"}
clr <- rgb(0, 0, 255, alpha = 50, maxColorValue = 255)
## time seq
t_idx_f <- seq(yr_frst, length.out = n_yrs+n_fore)
## plot
yp_min <- min(run_size)
yp_max <- max(run_size)

png(file.path(figsdir, "Fig_2_popn_size.png"),
     height = 4, width = 7, units = "in", res = 500)

par(mai = c(1,1,0.1,0.1), omi = c(0,0,0,0))
plot(t_idx_f, run_size[3,], ylim = c(yp_min,yp_max), type = "n",
     log = "y", xaxt = "n", yaxt = "n", bty = "L",
     xlab = "Year", ylab = expression(Spawners~(10^3)),
     main = "", cex.lab = 1.2)
polygon(c(t_idx_f, rev(t_idx_f)), c(run_size[3,], rev(run_size[1,])),
        col = clr, border = NA)
lines(t_idx_f, run_size[2,], col = "blue3", lwd = 2)
points(cbind(t_idx_f, exp(ln_dat_esc) + dat_harv), pch = 16, cex = 1)
axis(1, at = seq(2000, 2015, 3))
axis(2, at = c(5000, 10000, 20000, 40000),
     labels = c(c(5, 10, 20, 40)), las = 1)

dev.off()
```

```{r, echo=FALSE}
cnt <- cnt + 1
```

\newpage

\begin{center}
\includegraphics[width=0.9\textwidth]{figures/Fig_2_popn_size}
\end{center}

`r fig_cap(cnt, caption = "Time series of the estimated total population size. The observed data are the points; the solid line is the median estimate of the true number of spawners and the shaded region indicates the 95% credible interval.", display = "full")`

\newpage

## Spawner-recruit relationship

```{r plot_S_R, fig.width = 7, fig.height = 4.5, fig.pos = "placeHere", fig.align="center"}
png(file.path(figsdir, "Fig_3_SR.png"),
     height = 4.5, width = 7, units = "in", res = 500)

layout(matrix(c(1,1,2,3),2,2),c(3,2),c(1,1))
xoffSet <- 0.05
yoffSet <- 0.03

## colors for plotting
clr <- rgb(100, 0, 200,
           alpha = seq(200, 100,
                       length.out = age_max-age_min+n_fore),
           maxColorValue = 255)

dd <- 3000
yM <- Re2prec(max(r_dat), "ceiling", dd)
xM <- Re2prec(max(s_dat), "ceiling", dd)

## empty plot space for spawner-recruit relationships
par(mai = c(0.8,0.8,0.1,0.1), omi = c(0,0,0,0))
plot(s_dat[2,], r_dat[2,], xlim = c(0,xM), ylim = c(0,yM), type = "n",
     xaxs = "i", yaxs = "i", cex.lab = 1.2,
     xlab = expression(Spawners~(10^3)),
     ylab = expression(Recruits~(10^3)),
     xaxt = "n", yaxt = "n", bty="L")
axis(1, at = seq(0,xM,dd*2), labels = seq(0,xM,dd*2)/1000)
axis(2, at = seq(0,yM,dd*2), labels = seq(0,yM,dd*2)/1000, las=1)
for(i in 1:length(alphaCI)) {
  lines(alphaCI[i] * seq(0,xM) * exp(-bb * seq(0,xM)),
        lwd = 2, col = ifelse(i==2, "black", "darkgray"))
}
abline(a = 0,b = 1,lty = "dashed")

## add S-R estimates and medians
nCB <- n_yrs-age_max
## years with complete returns
points(s_dat[2, 1:nCB], r_dat[2, 1:nCB],
       xlim = c(0,xM), ylim = c(0,yM),
       pch = 16, col = "blue3")
segments(s_dat[2, 1:nCB], r_dat[1, 1:nCB],
         s_dat[2, 1:nCB], r_dat[3, 1:nCB],
         col = "blue3")
segments(s_dat[1, 1:nCB], r_dat[2, 1:nCB],
         s_dat[3, 1:nCB], r_dat[2, 1:nCB],
         col = "blue3")
nTB <- dim(s_dat)[2]
## years with incomplete returns
segments(s_dat[2, (nCB+1):nTB], r_dat[1, (nCB+1):nTB],
         s_dat[2, (nCB+1):nTB], r_dat[3, (nCB+1):nTB],
         col = clr)
segments(s_dat[1, (nCB+1):nTB], r_dat[2, (nCB+1):nTB],
         s_dat[3, (nCB+1):nTB], r_dat[2, (nCB+1):nTB],
         col = clr)
points(s_dat[2, (nCB+1):nTB],r_dat[2, (nCB+1):nTB],
       xlim = c(0,xM), ylim = c(0,yM),
       pch = 16, col = clr)
text(x = par()$usr[1] + diff(par()$usr[1:2]) * xoffSet,
     y = par()$usr[4] - diff(par()$usr[3:4]) * yoffSet,
     "(a)")

## posterior for alpha
clr <- rgb(0, 0, 255, alpha = 50, maxColorValue = 255)
par(mai = c(0.8,0.4,0.3,0.1))
hist(alpha_est, freq = FALSE, breaks = seq(0, Re2prec(max(alpha_est), "ceiling", 0.5), 0.2),
     col = clr, border = "blue3",
     xlab = "", ylab = "", main = "", cex.lab = 1.2, yaxt = "n")
aHt <- (par()$usr[4]-par()$usr[3])/12
arrows(alphaCI, par()$usr[3], alphaCI,par()$usr[3]-aHt,
       code = 1, length = 0.05, xpd = NA, col = "blue3", lwd = 1.5)
mtext(expression(Instrinsic~productivity~(alpha)), 1, line = 3, cex = 1)
text(x = par()$usr[1],
     y = par()$usr[4] * 1.05,
     "(b)", xpd=NA)

## posterior for K
par(mai = c(0.8,0.4,0.3,0.1))
brks <- seq(Re2prec(min(K_est), "floor"),
            Re2prec(max(K_est), "ceiling", 2))
hist(K_est, freq = FALSE, breaks = brks, col = clr, border = "blue3",
     xlab = "", xaxt = "n", yaxt = "n",
     main = "", ylab = "", cex.lab = 1.2)
axis(1, at = seq(Re2prec(min(K_est), "floor"),
                 Re2prec(max(K_est), "ceiling"),
                 2))
aHt <- (par()$usr[4] - par()$usr[3]) / 12
arrows(K_est_CI, par()$usr[3], K_est_CI,par()$usr[3]-aHt,
       code = 1, length = 0.05, xpd = NA, col = "blue3", lwd = 1.5)
mtext(expression(paste("Carrying capacity (",italic(K),", ",10^3,")")),
      side = 1, line = 3, cex = 1)
text(x = par()$usr[1], 
     y = par()$usr[4] * 1.05,
     "(c)", xpd=NA)

dev.off()
```

```{r, echo=FALSE}
cnt <- cnt + 1
```

\newpage

\begin{center}
\includegraphics[width=0.9\textwidth]{figures/Fig_3_SR}
\end{center}

`r fig_cap(cnt, caption = "Relationship between the number of spawning adults and their subsequent surviving offspring (recruits), assuming median values for the flow covariate (a); and the estimated posterior distributions for the intrinsic productivity (b) and carrying capacity (c). Points in (a) are medians of the posterior estimates; error bars indicate the 95% credible intervals. Blue points are for estimates with complete broods; purple points are for the most recent years with incomplete broods. The black line in (a) shows the expected relationship based on the median flow value over all years; gray lines indicate the 2.5$^\\text{th}$ and 97.5$^\\text{th}$ credible limits. Note that for plotting purposes only in (b) and (c), the density in the largest bin for each parameter contains counts for all values greater than or equal to it. Vertical arrows under the x-axes in (b) and (c) indicate the 2.5$^\\text{th}$, 50$^\\text{th}$, and 97.5$^\\text{th}$ percentiles of the posterior distribution.", display = "full")`

\newpage

## Recruits per spawner

```{r plot_R_per_S, fig.width = 7, fig.height = 4, fig.pos = "placeHere", fig.align="center"}
png(file.path(figsdir, "Fig_4_R_per_S.png"),
    height = 4, width = 7, units = "in", res = 500)

par(mai = c(1,1,0.1,0.1), omi = c(0,0,0,0))
clr <- rgb(0, 0, 255, alpha = 50, maxColorValue = 255)

## time sequence
t_idx_a <- seq(yr_frst, length.out = n_yrs-age_min+n_fore)
## plot data
p_dat <- mod_res[, grep("ln_RS", colnames(mod_res))]
p_dat <- apply(p_dat, 2, quantile, CI_vec)
yp_min <- min(p_dat)
yp_max <- max(p_dat)
## plot
plot(t_idx_a, p_dat[3,],
     type = "n",  bty = "L", xaxt = "n",
     ylim = c(yp_min,yp_max),
     xlab = "Brood year", ylab = "ln(R/S)", main = "",
     cex.lab = 1.2, las = 1)
abline(h = 0, lty = "dashed")
polygon(c(t_idx_a, rev(t_idx_a)), c(p_dat[3,], rev(p_dat[1,])),
        col = clr, border = NA)
lines(t_idx_a, p_dat[2,], col = "blue3", lwd = 2)
axis(1, at = seq(2000, 2015, 3))

dev.off()
```

```{r, echo=FALSE}
cnt <- cnt + 1
```

\newpage

\begin{center}
\includegraphics[width=0.9\textwidth]{figures/Fig_4_R_per_S}
\end{center}

`r fig_cap(cnt, caption = "Time series of the estimated recruits per spawner. The solid line is the median estimate and the shaded region indicates the 95% credible interval.", display = "full")`

\newpage

## Covariate effects

```{r plot_cov_effects, fig.width=7, fig.height=3.5, fig.pos="placeHere", warnings=FALSE, messages=FALSE, fig.align="center"}
clr <- rgb(0, 0, 255, alpha = 50, maxColorValue = 255)
xoffSet <- 0.05
yoffSet <- 0.03

c_est <- mod_res[,grep("gamma", colnames(mod_res))] * 1000
ylN <- Re2prec(min(c_est), "floor", 0.02)
ylM <- Re2prec(max(c_est), "ceiling", 0.02)
brks <- seq(ylN,ylM, length.out = diff(c(ylN,ylM))*400+1)

png(file.path(figsdir, "Fig_5_covar_effects.png"),
     height = 3.5, width = 7, units = "in", res = 500)

par(mfrow=c(1,2), mai=c(0.4,0.4,0.1,0.1), omi=c(0.5,0.5,0,0))

## plot covar ts
plot(dat_cvrs[, 1], dat_cvrs[, best_i+1]/1000,
     pch = 16, col = "blue3", type = "o",
     xlab = "", ylab = "", main = "", bty = "L", las = 1)
text(x = par()$usr[1] + diff(par()$usr[1:2]) * xoffSet,
     y = par()$usr[4] - diff(par()$usr[3:4]) * yoffSet,
     "(a)")
mtext(side = 2, expression(Flow~(10^3~cfs)), line = 3, cex = 1.2)
mtext(side = 1, "Brood year", line = 3, cex = 1.2)

## plot covar effect
hist(c_est,
     freq = FALSE, breaks = brks, col = clr, border =" blue3",
     xlab = "", yaxt = "n", main = "", ylab = "")
c_CI <- quantile(c_est,CI_vec)
aHt <- (par()$usr[4]-par()$usr[3])/20
arrows(c_CI, par()$usr[3]-0.005, c_CI, par()$usr[3] - aHt,
       code = 1,length = 0.05, xpd = NA, col = "blue3", lwd = 1.5)
abline(v = 0, lty = "dashed")
text(x = par()$usr[1] + diff(par()$usr[1:2]) * xoffSet,
     y = par()$usr[4] - diff(par()$usr[3:4]) * yoffSet,
     "(b)")
mtext(side = 1,expression(Effect~size~(10^{-3}~cfs)), line = 3, cex = 1.2)

dev.off()
```

```{r, echo=FALSE}
cnt <- cnt + 1
```

\newpage

\begin{center}
\includegraphics[width=0.9\textwidth]{figures/Fig_5_covar_effects}
\end{center}

`r fig_cap(cnt, caption = "Time series of the maximum of the 7-day mean flow during the months of February through April (a), and its estimated effect on population productivity (b). Small arrows under the histogram in (b) denote 2.5$^\\text{th}$, 50$^\\text{th}$, and 97.5$^\\text{th}$ percentiles of the posterior distribution.", display = "full")`

\newpage

## Process errors

```{r plot_proc_err, fig.width = 7, fig.height = 4, fig.pos = "placeHere", fig.align="center", results="asis"}
## time sequence
t_idx_a <- seq(yr_frst, length.out = n_yrs-age_min+n_fore)
## plot data
proc_err <- mod_res[, grep("res_ln_Rec", colnames(mod_res))]
proc_err <- apply(proc_err, 2, quantile, CI_vec)
yp_min <- min(proc_err)
yp_max <- max(proc_err)

png(file.path(figsdir, "Fig_6_proc_errs.png"),
     height = 3.5, width = 7, units = "in", res = 500)

phi_est <- mod_res[,grep("phi", colnames(mod_res))]
ylN <- Re2prec(min(phi_est), "floor", 0.02)
ylM <- Re2prec(max(phi_est), "ceiling", 0.02)
brks <- seq(ylN,ylM, length.out = diff(c(ylN,ylM))*20+1)

par(mfrow=c(1,2), mai=c(0.4,0.4,0.1,0.1), omi=c(0.5,0.5,0,0))

## plot proc errors
plot(t_idx_a, proc_err[3,],
     type = "n",  bty = "L", xaxt = "n", las = 1,
     ylim = c(yp_min,yp_max),
     xlab = "", ylab = "", main = "",
     cex.lab = 1.2)
abline(h = 0, lty = "dashed")
polygon(c(t_idx_a, rev(t_idx_a)), c(proc_err[3,], rev(proc_err[1,])),
        col = clr, border = NA)
lines(t_idx_a, proc_err[2,], col = "blue3", lwd = 2)
text(x = par()$usr[1] + diff(par()$usr[1:2]) * xoffSet,
     y = par()$usr[4] - diff(par()$usr[3:4]) * yoffSet,
     "(a)")
axis(1, at = seq(2000, 2015, 3))
mtext(side = 2, "Process error", line = 3, cex = 1.2)
mtext(side = 1, "Brood year", line = 3, cex = 1.2)

## plot covar effect
hist(phi_est,
     freq = FALSE, breaks = brks, col = clr, border =" blue3",
     xlab = "", yaxt = "n", main = "", ylab = "")
phi_CI <- quantile(phi_est,CI_vec)
aHt <- (par()$usr[4]-par()$usr[3])/20
arrows(phi_CI, par()$usr[3]-0.005, phi_CI, par()$usr[3] - aHt,
       code = 1,length = 0.05, xpd = NA, col = "blue3", lwd = 1.5)
abline(v = 0, lty = "dashed")
text(x = par()$usr[1] + diff(par()$usr[1:2]) * xoffSet,
     y = par()$usr[4] - diff(par()$usr[3:4]) * yoffSet,
     "(b)")
mtext(side = 1,expression(phi), line = 3, cex = 1.2)

dev.off()
```

```{r, echo=FALSE}
cnt <- cnt + 1
```

\newpage

\begin{center}
\includegraphics[width=0.9\textwidth]{figures/Fig_6_proc_errs}
\end{center}

`r fig_cap(cnt, caption = "Time series of the estimated process errors (a), which represent the population's productivity after accounting for the effects of density dependence and environmental covariates, and the posterior distribution of the autocorrelation parameter $\\phi$ (b). The solid line in (a) is the median estimate and the shaded region indicates the 95% credible interval. Small arrows under the histogram in (b) denote 2.5$^\\text{th}$, 50$^\\text{th}$, and 97.5$^\\text{th}$ percentiles of the posterior distribution.", display = "full")`


